\documentclass[a4paper, 11pt]{article}
\usepackage{marginnote}


\def\nterm {Winter}
\def\nyear {2023}
\def\nlecturer {Carlos Améndola}
\def\ncourse {Algebraic Statistics}

\input{header}
\usepackage{makeidx}
\usepackage[]{mdframed}
\usepackage{algpseudocode}
\usepackage{tikz-3dplot}

\makeindex

\newcommand{\PhantC}{\phantom{\colon}}%
\newcommand{\PhantSQ}{\phantom{\sqrt{\hspace{0.3ex}}}}%
\newcommand*{\vertbar}{\rule[-1ex]{0.5pt}{2.5ex}}
\newcommand*{\horzbar}{\rule[.5ex]{2.5ex}{0.5pt}}
\newcommand{\indep}{\perp \!\!\! \perp}

% https://tex.stackexchange.com/questions/63355/wrapping-cmidrule-in-a-macro
\ExplSyntaxOn
\makeatletter
\newcommand{\CMidRule}{\noalign\bgroup\@CMidRule{}}
\NewDocumentCommand{\@CMidRule}{
    m % Material to reinsert before cmidrule.
    O{0.0ex} % #1 = left adjust
    O{0.0ex} % #1 = right adjust
    m  %       #3 = columns to span
}{
    \peek_meaning_remove_ignore_spaces:NTF \CMidRule
      { \@CMidRule { #1 \cmidrule[\cmidrulewidth](l{#2}r{#3}){#4} } }
      { \egroup #1 \cmidrule[\cmidrulewidth](l{#2}r{#3}){#4} }
}
\makeatother
\ExplSyntaxOff


\begin{document}
\maketitle
\tableofcontents

\clearpage

\section{Introduction}

\subsection{Discrete Markov chain}

\marginnote{\small{Lec01}}

Let \( X_1, X_2, ..., X_m \) be a sequence of random variables on the same finite state space \( \Omega \). The joint probability distribution is \( P (X_1 = x_1, ..., X_m = x_m) 
\) with \( x_i \in \Omega \) such that \( \sum P(X_1 = x_1,...,X_n = x_n) = 1 \).

\begin{defi}[Markov chain]
  A sequence \( X_1, ..., X_n \) is called a \textbf{Markov chain} if the probability of the next state depends only on the current state; that is,
  \begin{align*}
    \mathbb P (X_{n+1} = x_{n+1} \mid X_1 = x_1, ..., X_n = x_n) = \mathbb P (X_{n+1} = x_{n+1} \mid X_n = x_n).
  \end{align*}
\end{defi}


\begin{mdframed}
  Motto in algebraic statistics: \textbf{Statistical models are semi-algebraic sets}.
\end{mdframed}

Here is an example of a statistical model that can be represented as a solution set of system of polynomial equations and inequalities.

\begin{prop}[Implicit Markov chain model]
  A vector \( p \in \mathbb R^8 \) is the probability distribution from a Markov chain model \( \mathcal{M} \) if and only if 
  \begin{itemize}
    \item \( p \in \Delta_7 \),
    \item \( p_{010}p_{111} - p_{011} p_{110} \),
    \item \( p_{000}p_{101} - p_{001}p_{100} \).
  \end{itemize}
  The second and third equality is obtained by taking the determinant of the matrix slices \( j=0 \) and \( j = 1 \):
  \begin{align*}
    \begin{bmatrix}
      p_{000} & p_{001} \\
      p_{100} & p_{101}
    \end{bmatrix} \quad \text{and} \quad 
    \begin{bmatrix}
      p_{010} & p_{011} \\
      p_{110} & p_{111}
    \end{bmatrix}.
  \end{align*}
\end{prop}


\begin{proof}[Proof]
  \( \implies \): 
  Assume we are given random variable \( X_1, X_2 \) and \( X_3 \) taking values in \( \Omega = \left\{ 0,1 \right\} \) and satisfying the Markov chain property. A joint probability distribution can be given as an \( \mathbb{R}^8 \) vector 
  \begin{align*}
    \mathbf p = \begin{bmatrix}
      p_{000} \\ p_{001} \\ \vdots \\ p_{111}
    \end{bmatrix} \in \Delta_7.
  \end{align*}
  By definition of conditional probability, we compute
  \begin{align*}
    &P(X_3 = k \mid X_1 = i, X_2 = j) = \frac{p_{ijk}}{p_{ij+}} \\
    &P(X_2 = j \mid X_1 = i) = \frac{p_{ij+}}{p_{i++}}.
  \end{align*}
  Using the Markov chain property, we obtain the equality 
  \begin{align*}
    \frac{p_{ijk}}{p_{ij+}} =  \frac{p_{ij+}}{p_{i++}}.
  \end{align*}
  Similarly, if for some different \( i \neq i' \) we obtain \( \frac{p_{i'jk}}{p_{i'j+}} =  \frac{p_{ij+}}{p_{i++}} \), and thus 
  \begin{align*}
    \frac{p_{ijk}}{p_{ij+}} = \frac{p_{i'jk}}{p_{i'j+}}.
  \end{align*}
  This yields \( p_{ijk} (p_{i'j0} + p_{i'j1}) = p_{i'jk} (p_{ij0} + p_{ij1}) \).
  \begin{itemize}
    \item Substitute \( i=j=k=1, i' = 0 \): \( p_{111}(p_{010} + p_{011}) = p_{011}(p_{110} + p_{111}) \). Thus, \( p_{111}p_{010} - p_{011}p_{110} = 0 \).
    \item Substitute \( i=j=k=0, i' = 1 \): \( p_{000}(p_{100} + p_{101}) = p_{100}(p_{000} + p_{001}) \). Thus, \( p_{000}p_{101} - p_{100}p_{001} = 0 \).
    \item Substitute \( i=j=1, k=0, i' = 0 \): \( p_{110}(p_{010} + p_{011}) = p_{010}(p_{110} + p_{111}) \). Thus, \( p_{110}p_{011} - p_{010}p_{111} = 0 \).
    \item And so on (we obtain the same equations)...
  \end{itemize}
\end{proof}

\begin{remark}
  A Markov chain model is an example of a \textbf{conditional independence model}.
\end{remark}


\begin{prop}[Parametrization of the Markov chain model]
  A parametrization of the Markov chain model is given by 
  \begin{align*}
    \varphi(\pi, \pmb \alpha) =  \begin{bmatrix} \vdots \\
      \pi_{j_i} \prod^m_{i=2}\alpha_{i, j_{i-1}, j_i}     \\ \vdots \end{bmatrix}  = \begin{bmatrix}
      p_{000} \\ p_{001} \\ \vdots \\ p_{111}
    \end{bmatrix} = \mathbf p \in \Delta_7
  \end{align*}
  where \( \pi_{j_i} = P(X_1 = j_i) \) and \( \alpha_{i,j_{i-1}, j_i} = P(X_i = j_{i} \mid X_{i-1} = j_{i-1}) \).
\end{prop}

\begin{eg}
Let us consider the case with three random variables \( X_1, X_2 \) and \( X_3 \). We define \( \alpha_{ij} = \alpha_{1ij} \) and \( \beta_{jk} = \alpha_{2jk} \). Here are all the parameters 
\begin{align*}
  \pmb \pi = (\pi_0, \pi_1), \pmb \alpha = \begin{bmatrix}
    \alpha_{00} & \alpha_{01} \\ \alpha_{10} & \alpha_{11}
  \end{bmatrix}, \pmb \beta = \begin{bmatrix}
  \beta_{00} & \beta_{01} \\ \beta_{10} & \beta_{11} \end{bmatrix}.
\end{align*}
Note that the components of \( \pi \) sum to one as well as the rows of \( \alpha \) and \( \beta \). Thus, we have five \emph{independent variables} \( \pi_0, \alpha_{00}, \alpha_{10}, \beta_{00} \) and \( \beta_{10} \). We later say that \( \mathrm{dim}(\mathcal{M}) = 5 \).
\end{eg}


\subsection{Maximum likelihood}

Assume we observe the following data \( D \):
\begin{align*}
  000, 010, 110, 000, 101, 110, 100, 010, 110, 111, 000, 000, 010.
\end{align*}
The count vector reads \( \mathbf u = (4,0,3,0,1,1,3,1) \). The likelihood function is
\begin{align*}
  L_{u}(p) = {n \choose \mathbf u} \prod_{i,j,k}p_{ijk}^{u_{ijk}}
\end{align*}
where \( {n \choose \mathbf u} = \frac{n!}{\prod_{i,j,k} u_{ijk}!} \) is the multinomial coefficient. We want to maximize \( L_u \). When we use the implicit model, we obtain a \emph{constrained optimization problem} 
\begin{align*}
  \argmax_{\mathbf p \in \Delta_7}\left \{\prod_{ijk}{p_{ijk}}^{u_{ijk}}\right \} \text{ subject to } \begin{cases}
    p_{010}p_{111} - p_{011} p_{110}  \\
    p_{000}p_{101} - p_{001}p_{100} 
  \end{cases}
\end{align*}
An \emph{unconstrained optimization problem} is given by the parametrized Markov model (although it is not quite unconstrained): optimize 
\begin{align*}
  \argmax_{\pmb \pi, \pmb \alpha, \pmb \beta} \left\{ \prod_{ijk} (\pi_i \alpha_{ij} \beta_{jk})^{u_{ijk}} \right\}
\end{align*}
with constraints that \( \pmb \pi, \pmb \alpha, \pmb \beta \) are probability distributions.

A typical strategy is to compute the logarithm of \( L_D \), equal the partial derivatives of that function to zero and solve the equations. The equations that we set to zero are called \textbf{score equations} or \textbf{critical equations}.


We define the log-likelihood function 
\begin{align*}
  l(\pi, \alpha, \beta \mid u) = \sum_{ijk}u_{ijk} ( \log(\pi_i) + \log(\alpha_{ij}) + \log(\beta_{jk})) 
\end{align*}
Computing the derivative
\begin{align*}
  \frac{\partial l}{\partial \pi_0} = \frac{u_{0++} (1 - \pi_0)}{1} = u_{1++}\pi_0
\end{align*}
Hence \( \hat \pi_0 = \frac{u_{0++}}{u_{+++}} \) where \( u_{+++} = N \) the total sample size. Other solutions \( \hat \alpha_{10} = \frac{u_{10+}}{u_{1++}} \), \( \hat \beta_{01} = \frac{u_{+01}}{u_{+0+}} \).

To make a prediction we compute \( \hat p_{101} = \hat \pi_1 \hat \alpha_{10} \hat \beta_{01} \) where \( \hat \pi_1 = \frac{u_{1++}}{u_{+++}} \). The estimator is then \( \hat p_{101} = \frac{u_{10+} u_{+01}}{u_{+++}u_{+0+}} \). In this way 
\begin{align*}
  \hat \pi_i = \frac{u_{i++}}{u_{+++}}, \hat \alpha_{ij} = \frac{u_{ij+}}{u_{i++}}, \hat \beta_{ij} = \frac{u_{+jk}}{u_{+j+}}, \hat p_{ijk} = \frac{u_{ij+} u_{+jk}}{u_{+++}u_{+j+}}.
\end{align*}

A closed form rational solution \( \hat{\mathbf p} \) exists for our Markov chain model 
\begin{align*}
  \hat p_{ijk} = \frac{u_{ij+} u_{+jk}}{u_{+++}u_{+j+}}.
\end{align*}

In general, score equations rarely have simple closed solutions; for instance the solution need not be unique.


\marginnote{\small{Lec02, 04/17/23}}

\begin{defi}[Homogeneous Markov chain]
  A \textbf{homogeneous Markov chain} is a Markov chain where the transition matrices stay the same for each step, i.e. \( \alpha = \beta \).
\end{defi}

\subsection{Hypothesis testing}
Having observed data, we found a distribution \( \hat p \in \mathcal{M} \) in the model that `best' explains the data. 

\begin{mdframed}
  How well does the model fit the data?\\
  Do we think that the data was generated by some distribution in the model?
\end{mdframed}

Let \( p \) be the true generating distribution of the count \( u \). Hypothesis test:  \(H_0: p \in \mathcal{M} \) vs \( H_1: p \notin \mathcal{M} \). (In practice, for example clinical tests, one often wants that \( H_0 \) does not occur. One wants to reject \( H_0 \). One wants to have enough evidence to reject \( H_0 \). Only take \( H_1 \) if there is enough evidence.). 

Idea: suppose \( p \in \mathcal{M} \) and we observe \( u \). Consider other outcomes \( V \) that could have been generated by \( p \). The question is: what proportion of such data is more likely to have occurred than \( u \)? If this proportion is large, then either
\begin{enumerate}
  \item hypothesis \( H_0 \) is false and \( H_1 \) holds 
  \item we were unlucky (a rare event happened)
\end{enumerate}

For example, we throw a dive and we assume that the dice is fair. Assume we observe that nine out of ten times it was heads. If we assume that \( H_0: \)the dice is fair, that count would be very rare.

No way to know if we were unlucky or \( H_0 \) is wrong. In practive, if probability is small enough \( < 5 \% \), then one says there is enough significant evidence to reject \( H_0 \). Otherwise, one says there is not enough evidence to reject.

Problem: probability of observing data \( v \) depends on (unknown) p.

Key insight from Fisher: Fisher's exact test: for some models the likelihood function does not depend on exact data but only through a (non-injective) function of the data. This function is called sufficient statistic. Back to \( m=3 \) with parameters \( \pi, \alpha, \beta \).
\begin{align*}
  P (\mathbf u \mid \pmb \pi, \pmb \alpha,\pmb \beta)  &=  {n \choose u} \prod_{ijk} (\pi_i \alpha_{ij} \beta_{jk})^{u_{ijk}} \\ 
  &= {n \choose u} \prod_{ijk} (\pi_i \alpha_{ij})^{u_{ij+}} \prod \beta_{jk}^{u_{+jk}}
\end{align*}
The quantities \( u_{ij+}, u_{+jk} \) are the \emph{sufficient statistics}.

Fishers ida: compare observed data \( u \) with other data \( v \). The definition of a Markov chain is 

\clearpage
\include{chapters/probability-primer}


\clearpage
\include{chapters/algebra-primer}



\clearpage
\include{chapters/conditional-independence}

\clearpage
\section{Statistics Primer}

\marginnote{\small{Lec13, 27.11.2023}}

\begin{defi}[Statistical model]
  A \textbf{statistical model} \( \mathcal{M} \) is a collection of probability distributions or density functions. A \textbf{parametric statistical model} \( \mathcal{M}_{\Theta} \) is a statistical model of the form \( \mathcal{M}_{\Theta} = \left\{ p_\theta : \theta \in \Theta \right\} \), where \( \Theta \subset \mathbb R^d \) is a finite-dimensional parameter space.
\end{defi}

\begin{defi}[Identifiable Model]
  A statistical model is called \textbf{identifiable} if \( p_\theta = p_{\theta '} \) implies \( \theta = \theta' \).
\end{defi}

\begin{eg}[Binomial random variable model]
  Let \( n \in \mathbb{N} \). For \( \theta \in \Theta = [0,1] \) we define \( p_\theta(k) = {n \choose k} \theta^k (1 - \theta)^{n - k} \). A statistical model is given by \( M_\theta = \left\{ p_\theta : \theta \in \Theta \right\}  \).
\end{eg}

\begin{eg}[Multivariate normal random vector]
  Let \( X \in \mathbb{R}^m \) be a real random vector. Let \( \Theta = \mathbb{R}^m \times \mathrm{PD}_m \). Define \( \mathcal{M}_\Theta = \left\{ p_\theta (x) : \theta \in \Theta \right\} \) where 
  \begin{align*}
    p_\theta = (2\pi)^{-\frac{m}{2}} |\Sigma|^{-\frac{1}{2}}\exp{\left( -\frac{1}{2} (x - \mu)^T \Sigma^{-1}(x - \mu) \right)}.
  \end{align*}
\end{eg}

\begin{defi}[Implicit statistical models]
  \textbf{Implicit statistical models} are given by constraints on the probability distributions or densities.
\end{defi}

\begin{eg}[Implicit binomial statistical model]
  The binomial model can be implicitly given by the intersection of \( \Delta_n \) with the \( 2 \times 2 \)-minors of 
  \begin{align*}
    \begin{bmatrix}
      \dfrac{x_0}{{n \choose 0}} & \dfrac{x_1}{{n \choose 1}} & \dfrac{x_2}{{n \choose 2}} & \dots & \dfrac{x_{n-1}}{{n \choose n-1}} \\[1.5em]
      \dfrac{x_1}{{n \choose 1}} & \dfrac{x_2}{{n \choose 2}} & \dfrac{x_3}{{n \choose 3}} & \dots & \dfrac{x_{n}}{{n \choose n}}
    \end{bmatrix}.
  \end{align*}
\end{eg}

Here is another implicit model.

\begin{eg}[Independent random variables]
  Let \( X,Y \) be two discrete random variables with \( X \in [r] \) and \( Y \in [c] \). Let \( \mathcal{R} = [r] \times [c] \) and \( \Delta_{\mathcal{R}} = \Delta_{n} \) where \( n = |\mathcal{R}| - 1 \). The \textbf{model of independence} \( \mathcal{M}_{X \indep Y} \) is defined by the constraints
  \begin{align*}
    \mathcal{M}_{X \indep Y} &= \Delta_{\mathcal{R}} \cap \{ \mathbf p \in \mathbb R^{r \times c} : \mathbb P(X = i, Y = j) = \mathbb P(X = i) \mathbb P(Y = j), \; \forall i \in [r], j \in [c]\} \\
    &=  \Delta_{\mathcal{R}} \cap \{ \mathbf p \in \mathbb R^{r \times c} : \mathrm{rank}(\mathbf p) = 1\}.
  \end{align*}
  This is an \emph{implicit} representation of \( \mathcal{M}_{X \indep Y} \). For a \emph{parametric} representation we define \( \theta = (\alpha, \beta) \in \Delta_{r - 1} \times \Delta_{c - 1} \) with \( \mathbb P_{\theta}(X_1 = i, X_2 = j) = \alpha_{i} \beta_{j} \).
\end{eg}

\begin{eg}[Main example of implicit statistical models]
  Consider a set of conditional independence statements \( \mathcal{C}  = \left\{ A_1 \indep B_1 \mid C_1, A_2 \indep B_2 \mid C_2, \dots \right\}\). If \( \mathbf X \) is a discrete random vector, then the discrete conditional independence model is implicit:
  \begin{align*}
    \mathcal{M}_{\mathcal{C}} = V\left(\sum_{\mathcal{C}} I_{A \indep B \mid C}\right) \cap I_{\Delta_{\mathcal{R}}}.
  \end{align*}
  If \( \mathbf X \sim \mathcal{N}_m(\mu, \Sigma) \) is a Gaussian vector, then the Gaussian conditional independence model is also implicit
  \begin{align*}
    \mathcal{M}_{\mathcal{C}} = V\left(\sum_{\mathcal{C}} I_{A \indep B \mid C}\right) \cap \mathrm{PD}_{m}.
  \end{align*}
\end{eg}

\begin{defi}[Statistic]
  Let \( \mathbf X \in \mathcal{X} \) be a random vector with state space \( \mathcal{X} \).
  A \textbf{statistic} is a function from the state space \( \mathcal{X} \) to some set.
\end{defi}

Examples are 
\begin{itemize}
  \item \( \bar X_n = \frac{1}{n} \sum^n_{i=1}X_i \),
  \item \( S^2_n = \frac{1}{n-1} \sum^n_{i=1} (X_i - \bar X_n)^2 \),
  \item \( T = \max\left\{ X_1,X_2, ..., X_n \right\} \),
  \item \( T' = 5 \).
\end{itemize}

We informally describe what a sufficient statistic is. We say \( T \) is a \textbf{sufficient statistic} if the statistician who knows the value of \( T \) can do just as good a job of estimating the unknown parameter \( \theta \) as the statistician who knows the entire random sample.

\begin{defi}[Sufficient statistic]
  A statistic \( T \) is called \textbf{sufficient} if \( \mathbb P(X = x \mid T(X) = t, \theta) = \mathbb P(X = x \mid T(X) = t) \).
\end{defi}

\begin{thm}[Factorization theorem]
  Let \( \mathbf X \) be a random vector with joint density \( f(\mathbf x \mid \theta) \). A statistic \( T \) is sufficient if and only if the joint density \( f \) can be factored as follows
  \begin{align*}
    f(\mathbf x \mid \theta) = u(\mathbf x) v(T(\mathbf x), \theta)
  \end{align*}
  for non-negative functions \( u,v \).
\end{thm}

\begin{eg}[Sufficient statistic of the likelihood function of a multivariate normal distribution]
  Given samples \( X^{(1)},  \dots, X^{(N)} \) i.i.d. normal, the joint distribution is given by 
  \begin{align*}
    p_{\mu, \Sigma}( X) = \prod^N_{i=1} \frac{1}{(2\pi)^{\frac{m}{2}} |\Sigma|^\frac{1}{2}} \exp{\left( -\frac{1}{2} (X^{(i)} - \mu)^T \Sigma^{-1}(X^{(i)} - \mu) \right)}.
  \end{align*}
  A sufficient statistic is given by 
  \begin{align*}
    T( X) = \begin{bmatrix}
      \bar X \\
      \frac{1}{N} \sum^N_{i=1} (X^{(i)} - \bar X)(X^{(i)} - \bar X)^T
    \end{bmatrix}.
  \end{align*}
\end{eg}



\clearpage


\subsection{4.12}

Last topic: Parameter estimation. Given a sample \( X^{(1)}, X^{(2)}, ..., X^{(N)} \) iid \( \sim P_\theta \in \mathcal{M}_{\Theta} \) estimate the value of \( \theta \in \Theta \). We have an estimator \( \hat \theta (X^{(1)} ,  ... , X^{(N)}) \).

\begin{eg}
  Assume \( X^{(1)}, ..., X^{(N)}  \sim \mathrm{Ber}(\theta)\)  with \( \mathbb{E} [\bar X_N] = \theta \) unbiased, \( \hat \theta (X^{(1)}, ..., X^{(N)}) \coloneqq \bar X_N \).
\end{eg}

Methods of moment: let \( X = (X_1,..., X_m) \) be a random vector. Let \( \alpha \in \mathbb N^m \), the \( \alpha \)-th moment is defined to be \( \mu_\alpha = \mathbb E[X_1^{\alpha_1} X_2^{\alpha_2} ... X_m^{\alpha_m}] \). The order of moment is \( |\alpha| = \sum \alpha_i \).

Another example: Binomial models

\begin{eg}[Multidimensional gaussian]
  Assume \( X^{(1)}, ..., X^{(N)} \sim N_m ( \mu, \Sigma) \).
  \begin{itemize}
    \item \( M_1 \coloneqq \mathbb{E}[X] = \mu \in \mathbb R^m \).
    \item \( M_2 \coloneqq \mathbb{E}[XX^T] = \Sigma + \mu\mu^T \)
    \item \( M_3 \coloneqq \mathbb{E}[X \otimes X \otimes X] \)
  \end{itemize}
  The first two order moment suffice to solve the system.
  \begin{align*}
    \hat \mu_1 = \mu, \hat \mu_2 = \Sigma + \mu\mu^T
  \end{align*}
  and 
  \begin{align*}
    \hat \Sigma = \hat M_2 - \hat M_1 \hat M_1^T &= \frac{1}{N} \sum^N_{i=1} X^{(i)} (X^{(i)})^T - \bar X_N \bar X_N^T \\
    &= \frac{1}{N} \sum^N_{i=1} (X^{(i)} - \bar X)(X^{(i)} - \bar X)^T = \frac{N-1}{N}S_N^2 = \tilde S^2_N
  \end{align*}
\end{eg}

\textbf{Sample covariance: } \( S_N^2 = \frac{1}{N- 1} \sum^N_{i=1} (X^{(i)} - \bar X)^2 \). 

Given \( X^{(1)}, ..., X^{(N)} \) iid, \( \mathbb{E}[X^{(i)}] = \mu \) and \( \mathrm{Var}[X^{(i)}] = \sigma^2 \). Is the sample covariance unbiased, i.e. \( \mathbb{E}[S_N^2] = \sigma^2 \) and \( \mathbb{E}[\tilde S^2_N] = (1 - \frac{1}{N})\sigma^2 \).

\begin{proof}
  \begin{align*}
    \mathbb{E}[\sum^N_{i=1} (X^{(i)} - \bar x)^2] &= \sum^N_{i=1} \mathbb E [(X^{(i)} - \bar x)^2] = \sum^N_{i=1} \mathrm{Var}[X^{(i)} - \bar x] = \sum^N_{i=1} \mathrm{Var}[ (1 - \frac{1}{N})X^{(i)}  - \frac{1}{N}\sum_{j}^N X^{(j)}] \\
&= \sum^N_{i=1}((1 - \frac{1}{N})^2 \mathrm{Var}[X^{(i)}] + \frac{1}{N^2}\sum \mathrm{Var[X^{(j)}]}) \\
&= \sum (\frac{N - 1}{N})^2 \sigma^2 + \frac{N - 1}{N^2} \sigma^2 \\
...
  \end{align*}
\end{proof}

Gaussian mixture: \( f(x) = \lambda f_{\mu_1, \sigma_1^2}(x) + (1 - \lambda) f_{\mu_2,  \sigma_2^2} (x) \). Assume \( X_1 \sim N(\mu, \sigma_1^2) \) and \( X_2 \sim N(\mu_2, \sigma^2_2) \).

Assume \( X^{(1)}, ..., X^{(N)} \) iid \( \sim f \). Then \( m_1 = \lambda \mu_1 + ( 1 - \lambda) \mu_2 \) and \( m_2 = \lambda (\mu_1^2 + \sigma_1^2) + (1 - \lambda)(\mu_2^2 + \sigma_2^2) \), \( m_3 = \lambda(\mu_1^3 + 3 \mu_1 \sigma_1^2)  + ...\), \( m_4 = \lambda(\mu_1^4 + 6 \mu_1^2 \sigma_1^2 + 3 \sigma_1^4) + (1 - \lambda)... \) and \( m_5 = \lambda(\mu_1^5 + 10 \mu_1^3 \sigma_1^2 + 15 \mu_1 \sigma_1^4) + (1- \lambda) \).

Solve system for \( \lambda, \mu_1, \mu_2, \sigma_1^2, \sigma_2^2 \) and \( \hat m_1 = m_1, ..., \hat m_5 = m_5 \). Reduces to monis polynomial in \( p = \mu_1 \mu_2 \).

\begin{eg}
  Assume \( X^{(1)}, ..., X^{(N)} \sim \mathrm{Unif}(-\theta, \theta)\)  with \( \theta > 0 \). What is \( \hat \theta_{MOM} \)? So \( \bar x_\mu = \hat \mu_1 = \mu_1(\theta) = 0 \). This does not work.

  \( \hat \theta_{MOM} = \sqrt{3 \hat \mu_2} \).
\end{eg}

\begin{eg}
  \( (x^{(1)}, ..., x^{(5)}) = (0,2,-1,-1, 0)  \), \( \hat \mu_1 = 0 \) and \( \hat \mu_2 = \frac{1}{5} (0^2 + 2^2 + 1 + 1 + 0^2) = \frac{6}{5} \). So \( \hat \theta_{MOM}(x^{(1)}, ..., x^{(5)} = \sqrt{\frac{18}{5}}) \). THis is smaller than \( \sqrt{\frac{20}{5}} = 2 \). Not a good estimator.
\end{eg}

\subsection{Maximum likelihood estimation}

We recall that \( \frac{\partial \mathbf x^T \mathbf A \mathbf x}{\partial \mathbf x} = \mathbf x^T (\mathbf A + \mathbf A^T) \), \( \frac{\partial \mathrm{tr}(\mathbf A \mathbf B)}{\partial \mathbf A} = \mathbf B^T \), and \( \frac{\partial \log(\det \mathbf A)}{\partial \mathbf A} = (\mathbf A^{-1})^T \).

\begin{prop}[Multivariate normal model]\label{mle:multivariate-normal}
  For a multivariate normal random vector, the maximum likelihood estimator of \( \mu \) and \( \Sigma \) are given by 
  \begin{align*}
    \hat \mu = \frac{1}{N} \sum_{i=1}^N X^{(i)}, \quad \text{and} \quad \hat \Sigma = \frac{1}{N}(X^{(i)} - \hat \mu)(X^{(i)} - \hat \mu)^T.
  \end{align*}
\end{prop}

\begin{proof}
  The log-likelihood function is given by 
  \begin{align*}
    \ell_{X}( \mu, \Sigma) = -\frac{1}{2} \sum_{i=1}^N \left( m\log(2\pi) +  \log(\mathrm{det}(\Sigma)) + (X^{(i)} - \mu)^T \Sigma^{-1} (X^{(i)} - \mu) \right).
  \end{align*}
  Write \( X^{(i)} - \bar X + \bar X - \mu \) for \( X^{(i)} - \mu \), and we see that the log-likelihood function becomes 
  \begin{align*}
    \ell_{X}(\mu, \Sigma) = -\frac{mN}{2}\log(2 \pi) - \frac{N}{2} \log\det(\Sigma) -\frac{N}{2} \mathrm{tr}(\Sigma^{-1}S) - \frac{N}{2} (\bar X - \mu)^T \Sigma^{-1} (\bar X - \mu).
  \end{align*}
  Using the derivative rules we see that 
  \begin{align*}
    \hat \mu = \bar X \quad \text{and} \quad \hat \Sigma = S = \frac{1}{N} \sum (X^{(i)} - \hat \mu)(X^{(i)} - \hat \mu)^T.
  \end{align*}
\end{proof}


\clearpage
\include{chapters/exp-families}

\clearpage
\include{chapters/likelihood-inference}

\clearpage
\include{chapters/fishers-exact-test}


\clearpage
\section{Review}

\begin{itemize}
  \item What is the saturation ideal?
  \item What is the marginal density? 
  \item What is the conditional density?
  \item What is conditional independence?
  \item Pearl's conditional independence axioms
  \item What is the intersection axiom?
  \item Rank criterion for independence
  \item Algebraic criterion for conditional independence of discrete random vectors
  \item Conditional independence ideal of discrete random vectors
  \item Conditional independence model of discrete random vectors
  \item Rank criterion for conditional independence of Gaussian random vectors
  \item Algebraic criterion for conditional independence of Gaussian random vectors
  \item Conditional independence ideal of Gaussian random vectors
  \item Compute the conditional independence model of Gaussian random vectors for \( \mathcal{C} = \left\{ 1 \indep 3, 1 \indep 3 \mid 2 \right\} \). We should get \( 1 \indep \{2,3\} \) or \( 3 \indep \{1,2\} \).
  \item Prove that \( (1 \indep 2 \mid 3) \) and \( (2 \indep 3) \) imply \( 2 \indep (1,3) \).
  \item Prove the closure theorem
  \item How to compute the \( j \)-th elimination ideal? Use the Elimination Theorem, see Theorem \ref{thm:elimination-theorem}.
  \item How do you compute the closed image of a variety under a projection? What is \( \overline{\pi(V(I))} \)?
  \item Implicitization of binomial model with two trials
  \item Implicitization of general binomial model 
  \item What is a statistical model?
  \item What is a sufficient statistic?
  \item Factorization theorem
  \item Log likelihoo function of a multivariate normal distribution
  \item Sufficient statistic of the likelihood function of a multivariate normal distribution
  \item MLE of Gaussian model
  \item Exponential family definition
  \item Exponential form of binomial model
  \item Exponential form of univariate normal model
  \item Log-affine linear models
  \item What is a toric ideal?
  \item Toric ideals and binomial ideals
  \item Sufficient statistic and natural parameter of Gaussian model
  \item Inverse linear space
  \item Gaussian linear concentration model
  \item Concentration matrix and conditional independence
  \item Assume concentration matrix \( K \) has elements \( k_{12} = k_{13} = 0 \). What does this mean? What is the ideal?
  \item Maximum Likelihood degree definition
  \item Log-likelihood function of a discrete model 
  \item Score equations of a discrete model
  \item Compute MLE for binomial model with \( r \) trials given samples \( X^{(1)}, ..., X^{(n)} \)
  \item What is the ML degree of the twisted cube?
  \item What is the ML degree of the binomial model with two trials?
  \item What is the ML degree of the independence model?
  \item What is the ML degree of the full Gaussian model?
  \item What is the ML degree when covariance is ID matrix and mean is from a nodal cubic \( t^2 - 1,t(t^2 - 1) \)?
  \item MLE of restricted covariance Gaussian model with centered / unrestricted mean
  \item MLE of full independence Gaussian model and centered
  \item Likelihood in projective space 
  \item Likelihood degree of projective variety
  \item Lagrange multipliers
  \item Score equations with lagrange multipliers
  \item Example: discrete with \( r = 3 \) and \( V = ( p_1^2 - 
  4p_0p_2 ) \)
  \item MLE of independence model 2x2
  \item MLE of binomial model (2 trials)?
  \item June Huh Theorem
  \item June Huh applied on binomial model with two trials
  \item June Huh and independence model
  \item Birchs Theorem
  \item When does the MLE exist for log affine linear models?
  \item Define the Gaussian linear concentration model
  \item MLE of Gaussian linear concentration model 
  \item Gaussian graphical model
  \item MLE of graphical model
  \item Example: Compute the MLE of the Gaussian graphical model with \( 1 \indep 2 \mid 3 \)
  \item Chi square statistic
  \item Asymptotic \( p \)-value, how does it work?
  \item Probability of observing a contingency table; what is the problem?
  \item Definition of hypergeometric distribution
  \item Conditional distribution of observing a contingency table for \( r=c=2 \) and general case
  \item exact \( p \)-value
  \item explicit formula of the \( p \)-value
  \item Probability of observing a contingency table of a log-affine linear model
  \item What is a fiber?
  \item Conditional probability of observing a contingency table of a log-affine linear model + Independence Proof
  \item Exact conditional \( p \)-value of a log-affine linear model
  \item Markov basis
  \item Fundamental theorem of Markov basis
  \item Example: 2x2 independence model
\end{itemize}

\printindex

\end{document}
